title: "Cook County Property Tax Revenue Analysis - Multi-Unit Focus"
author: "Cook County Potential"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load required packages
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("knitr")) install.packages("knitr")
if (!require("scales")) install.packages("scales")
if (!require("ggplot2")) install.packages("ggplot2")

library(tidyverse)
library(knitr)
library(scales)
library(ggplot2)

# Helper functions for formatting
format_currency <- function(x) {
  scales::dollar(x, accuracy = 1)
}

format_number <- function(x) {
  scales::comma(x, accuracy = 1)
}

format_percent <- function(x) {
  scales::percent(x/100, accuracy = 0.1)
}

# Modern color palette
modern_palette <- c("#2E86AB", "#A23B72", "#F18F01")

# Helper function to simplify property types
simplify_property_type <- function(type) {
  case_when(
    type == "Condo" ~ "Condo",
    type == "Large Multi-Family (7+ units)" ~ "Large Multi-Family",
    type == "Small Multi-Family (2-6 units)" ~ "Small Multi-Family",
    TRUE ~ type
  )
}
```

## Introduction

This analysis examines property tax revenue from multi-unit residential development in Cook County, with a focus on Lake View Township. We analyze current tax patterns across different housing types and project potential revenue from new mixed-housing development.

## Methodology

### Data Sources
- Cook County Assessor's Office property records
- 2021-2022 property tax bills
- Township-level assessment data
- Market analysis of new construction prices
- Developer pro forma analysis

### Analysis Parameters
- **Property Values**: Based on current market rates for new construction
  - Condominiums: $575,000 per unit
  - Mid-market multi-family: $275,000-350,000 per unit
  - Luxury multi-family: $400,000+ per unit
- **Tax Rates**: Using current Cook County effective tax rates
- **Geographic Scope**: Lake View Township focus
- **Time Period**: Based on 2021-2022 baseline data

## Current Housing Stock Analysis

### Tax Bills and Assessed Values by Housing Type

```{r housing_analysis}
# Load the data
summary_data <- read.csv("property_tax_summary.csv")
township_data <- read.csv("property_tax_by_township.csv")
new_construction_rates <- read.csv("new_construction_tax_rates.csv")

# Create detailed summary statistics
detailed_summary <- summary_data %>%
  filter(year == 2022,
         property_type %in% c("Condo", "Large Multi-Family (7+ units)", "Small Multi-Family (2-6 units)")) %>%
  group_by(property_type) %>%
  summarize(
    avg_tax = mean(avg_tax_per_unit),
    median_tax = median(avg_tax_per_unit),
    avg_value = mean(avg_value_per_unit),
    median_value = median(avg_value_per_unit),
    effective_tax_rate = mean(effective_tax_rate)
  ) %>%
  mutate(property_type = simplify_property_type(property_type))

# Create summary table
detailed_summary %>%
  mutate(
    avg_tax = format_currency(avg_tax),
    median_tax = format_currency(median_tax),
    avg_value = format_currency(avg_value),
    median_value = format_currency(median_value),
    effective_tax_rate = format_percent(effective_tax_rate)
  ) %>%
  kable(col.names = c("Property Type", "Avg Tax/Unit", "Median Tax/Unit", 
                      "Avg Value/Unit", "Median Value/Unit", "Effective Tax Rate"),
        caption = "2022 Property Tax and Value Summary by Housing Type",
        align = c("l", "r", "r", "r", "r", "r"))
```

### Lake View Township Mixed Housing Development Scenario

This scenario represents a balanced approach to new development in Lake View Township, incorporating a mix of housing types to serve diverse needs while generating substantial tax revenue.

```{r lakeview_mixed_housing}
# Define Lake View mixed housing scenario
lakeview_mixed <- tribble(
  ~property_type, ~bedroom_type, ~units, ~value_per_unit,
  "Condo", "beds_2", 500, 575000,
  "Small Multi-Family (2-6 units)", "mixed_1_to_3", 2000, 312500,
  "Large Multi-Family (7+ units)", "mixed_studio_to_3", 1000, 400000
)

# Calculate effective tax rate (using 2% as example)
effective_tax_rate <- 0.02

# Calculate revenue
lakeview_results <- lakeview_mixed %>%
  mutate(
    total_value = units * value_per_unit,
    tax_revenue = total_value * effective_tax_rate,
    revenue_per_unit = tax_revenue / units,
    property_type = simplify_property_type(property_type)
  )

# Create detailed table
lakeview_results %>%
  select(property_type, units, value_per_unit, total_value, tax_revenue, revenue_per_unit) %>%
  mutate(across(where(is.numeric), ~ format_currency(.))) %>%
  arrange(desc(tax_revenue)) %>%
  kable(col.names = c("Property Type", "Units", "Value per Unit", "Total Value", 
                      "Annual Tax Revenue", "Revenue per Unit"),
        caption = "Lake View Township Mixed Housing Development - 3,500 Total Units",
        align = c("l", "r", "r", "r", "r", "r"))

# Create visualization
ggplot(lakeview_results, 
       aes(x = property_type, 
           y = tax_revenue/1e6, 
           fill = property_type)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = scales::dollar(tax_revenue/1e6, accuracy = 0.1, suffix = "M")),
            position = position_stack(vjust = 0.5),
            color = "white",
            size = 4) +
  scale_y_continuous(labels = scales::dollar_format(suffix = "M"),
                    expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = modern_palette) +
  labs(
    title = "Projected Annual Tax Revenue by Property Type - Lake View Township",
    subtitle = "Mixed Housing Development Scenario",
    x = "Property Type",
    y = "Tax Revenue (Millions)",
    fill = "Property Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray50"),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    panel.grid.major.x = element_blank()
  )
```

## Key Findings

### Current Housing Stock Insights
- Condos show the highest median property values but varied tax bills
- Large multi-family buildings demonstrate efficient tax revenue generation per unit
- Small multi-family properties provide a balanced mix of affordability and revenue

### Lake View Township Development Impact
- Mixed housing development could generate significant annual tax revenue:
  - Small multi-family units: $12.5M
  - Large multi-family units: $8.0M
  - Condos: $5.75M
  - Total projected annual revenue: $26.25M
- The scenario provides diverse housing options while maintaining strong revenue generation
- Development mix aligns with neighborhood character and market demands
